\chapter{Alternation} 
\label{chap:alts}

\input{alt1} % Introduction
\input{alt2} % Dining Philosophers, logging.
\input{alt3} % Avoiding deadlocks, With outports
\input{alt4} % quicksort bag of tasks example.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Summary}

In this chapter, we have studied \emph{alternation} constructs, whereby a
thread can try to communicate on any one of several different ports.  We
have seen several examples using alternation.  In particular, we saw how it
could be used to implement a bag of tasks with replacement. 

There is quite a lot of syntax involved with alternations: |alt| and |serve|;
in-port branches and out-port branches; guards; and indexed alternations.  You
should make sure you are familiar with the syntax.

There are two restrictions on the use of alternations: a port may not be
simultaneously feasible in two alternations; and both ports of a channel may
not be simultaneously feasible in alternations.  (Each restriction represents
a design choice, to make the implementation of alternation simpler and more
efficient.)

Along the way, we discussed how to use logging to help with debugging; later
we will use similar techniques for testing.  We also discussed how to identify
possible deadlocks as cycles of waiting, and how to avoid them.

As a caveat, you should note that alternations are fairly heavyweight
constructs.  Under the bonnet, the implementation has to do quite a lot of
work to identify whether any of its ports is ready for communication.  Don't
use an alternation unless there is a good reason to do so, i.e.,~you really
need a thread to be able to communicate on either of two channels.  Sometimes
it's easier and more efficient to force communications to happen in a
particular order, rather than allowing multiple different orders.

Sometimes a shared channel can produce the same effect as an \SCALA{alt}.  For
example, we could construct a variant of the dining philosophers where each
philosopher has channels \SCALA{pick} and \SCALA{drop} for each of its forks,
and where each fork has single \SCALA{pick} and \SCALA{drop} channels, on
which it can receive messages from either of the adjacent philosophers:
%
\begin{scala}
  def Fork(me: Int, pick: ??[Unit], drop: ??[Unit]) = thread("Fork"+me){
    repeat{ pick?(); drop?() }
  }
\end{scala}
%
%% Note that the \SCALA{pick} communication could be from either neighbouring
%% philosopher, so these need to be |ManyOne| channels.
Note that we need separate |pick| and |drop| channels, so that when one of the
philosophers is holding the fork, the other cannot pick it up.  However, such
a technique isn't always applicable. 

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Exercises}
 
\input{Exercises/changeMachine} % Change machine 

\input{Exercises/unboundedBuff} % Unbounded buffer 

\input{Exercises/diningPhils} % DP practical

\input{Exercises/adaptive} % Adaptive quadrature
