\chapter{Message Passing using Channels} 

\input{channels1} % Channels; multiples of 4; buffered channels
\input{channels2} % Closing channels; natural numbers
\input{usingChannels} % merge sort; order of actions

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% %%%%%

\section{Summary}

In this chapter we have seen the basics of message passing concurrency.  The
idea is straightforward: threads can send messages to one another.

We have arranged that threads have no shared variables.  This means that we
avoid any memory races, of the form that we saw in Chapter~\ref{chap:intro}.
In later chapters, we will relax this condition, and allow some shared
variables.  However, we will still need to ensure disciplined access to those
variables, and message passing will be one technique that we can use to
coordinate the threads.  

We have seen how to use message passing using the SCL library.  Threads
communicate via \emph{channels}, which comprise an \emph{in-port} and an
\emph{out-port}.  Threads can send messages at an out-port, which are received
at the corresponding in-port.  Channels can be either \emph{synchronous} or
\emph{asynchronous} (buffered).  A port can be shared between several threads.
Channels can be closed, for example to indicate the end of a stream of data;
and this can be used to signal the termination of a concurrent network.  

We have seen a number of examples illustrating message passing.  These have
mostly used fine-grained concurrency, where systems are built from very small
components.  The exercises at the end of this chapter include similar
examples.  Our aim in these examples has been to illustrate how concurrent
components can cooperate together.

We have also seen a technique for reasoning about concurrent programs, using
the ``happens-before'' relation ($\preceq$).

%%%%% 

\paragraph{Object-oriented versus thread-oriented programming}

An important part of programming is \emph{modularisation}: splitting a large
program up into manageable chunks.  The most common modularisation technique
is object orientation, where objects encapsulate some data and operations on
that data.  However, concurrency is also a form of modularisation, where each
thread holds some data, and communicates the results of operations on that
data with other threads.

In object-oriented systems, many objects can exist simultaneously, each with
its own state.  Normally, a single one is active at a time.  They communicate
by procedure calls, which pass data and control.

In message-passing concurrent systems, many threads can exist simultaneously,
each with its own state.  Often, all are active at the same time.  They
communicate by sending and receiving messages, which pass data.

The two techniques are complementary.  In later chapters, we will sometimes
encapsulate a thread inside an object.  And sometimes an individual thread
will use multiple objects, either as private objects or shared with other
threads. 

%%%%% 

\paragraph{Other approaches}

In this book, we use Scala together with the Scala Concurrency Library (SCL).
However, message-passing concurrency can also be used in a number of other
languages or with different concurrency primitives.

Message-passing concurrency started with the programming language {\sf occam}.
It was developed by INMOS in the early 1980s, targeting their transputer
microprocessors.

Concurrent Scala Objects (CSO) is a library of concurrency primitives, mostly
message-passing, developed by Bernard Sufrin.  SCL is heavily based on CSO,
and inherits much of the same syntax.  

Go\footnote{\url{https://golang.org/}} is a programming language created
by Google that uses message passing.  It is probably the most
mainstream such language.


Rust supports concurrency in a number of styles, including message
  passing\footnote{%
    \url{https://doc.rust-lang.org/book/ch16-02-message-passing.html}}. 
Erlang uses similar ideas, in a functional setting.

Message-passing libraries are available for a number of other languages.  For
example, JCSP\footnote{\url{http://www.cs.kent.ac.uk/projects/ofa/jcsp/}} is
an implementation  for Java.

% \item Eclectic
% CSP\footnote{\url{http://users.comlab.ox.ac.uk/bernard.sufrin/ECSP/ecsp.pdf}}
% is an experimental language based on CSP, and a precursor of CSO.

%% \item Communicating Haskell
%% Processes\footnote{\url{http://www.cs.kent.ac.uk/projects/ofa/chp/}} is a
%% Haskell library to support CSP-style communication.


% \item
% There are various experimental languages based on Python with CSP-style
% communication. 



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\exercises

\input{Exercises/boundedNats} % bounded version of NatsCircuit

\input{Exercises/hamming} % Hamming numbers

\input{Exercises/pipesort} % sorting using a pipeline.

\input{Exercises/quicksort}

\input{Exercises/sorting} % sorting using comparators


% More exercises: Bitonic part of sorting practical?
