\chapter{Clients and Servers}

%%%%%

\input{client-server1} % intro, resource allocation example, asynchrony
\input{client-server2} % testing; reply channels; buffering; 
\input{client-server3} % exchanger, filter channel

%\input{client-server4} % layering; conclusions

%%%%%

\section{Summary}

In this chapter we have studied the pattern of clients and servers: clients
make requests to a server, and the server responds.  

Typically, a server will allow different sorts of request.  These can be
supported by using a different channel for each sort, and using an alternation
in the server.  Alternatively, we could use a single channel, and use
different subtypes of the channel's data type to represent the different sorts
of request. 

Responses to requests (where needed) can be implemented in a couple of
different ways.  We could use a separate channel for each client.
Alternatively, we can arrange for the client to create a short-term reply
channel which it sends to the server, and for the server to respond on this
reply channel.  The latter approach is more flexible, so it is what we tend to
use. 

Sometimes a server is not immediately able to respond to a request; this is
the case in the resource allocation example, when no resource is available.
The normal approach in these circumstances is to store the request in a queue.
However, sometimes it is simpler for the server to refuse to communicate on
the relevant channel, so clients are blocked on that communication.  This
latter approach can also be useful in synchronisation objects, by restricting
communications with different sorts of clients to happen in a fixed order. 

Testing is an important consideration for concurrent objects.  We saw some
techniques in this chapter; and the same techniques are applicable regardless
of how the concurrent objects themselves are implemented.

Logging is often necessary for testing concurrent objects.  This is
particularly the case when the object maintains some state: we need to test
whether the results of operation executions are consistent with that state.
We will use this technique further when we consider concurrent datatypes in
the next chapter.

Logging is also often necessary when testing synchronisation objects.  We need
to identify which operation executions synchronise together, and for this we
need to know which overlap in time.  (The exchanger example was unusual in
this respect as we could identify which operations synchronised from their
parameters and results.)
 


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Exercises}

\input{Exercises/menWomenServer}

\input{Exercises/atomicBroadcast}

Possible \framebox{extra} exercises:
One family problem (2023 exam);
Same-arg synchronisation (2022 exam);
ABC problem (2021 exam).


