\chapter{Example: Numerical Integration}
\label{chap:trapezium}

\input{trapezium1} % up to testing
\input{trapezium2} % experiments
\input{trapezium3} % bag of tasks
\input{trapezium4} % experiments
\input{trapezium5} % encapsulation 

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Summary}

In this chapter we have studied a data parallel problem.  The approach to such
problems is to split the input data between different worker threads, and to
combine the results together.  Many compute-intensive problems fall into this
class, so it is a very useful pattern.  It  requires that there are
meaningful ways to split the data between workers, and to combine the results
efficiently.

We arranged for a controller thread to distribute tasks to workers, using two
different approaches.  In the first approach, the controller sent each worker
a single task.  (In fact, in this case we could have included the task as a
parameter of the |worker| function; but we wanted to illustrate this technique
of distributing tasks.)  In the second approach, the bag of tasks, the
controller distributed multiple tasks to workers; this helped to balance the
load between workers, overcoming differences in speed.

We tested each implementation against a corresponding sequential algorithm.
This is a very common technique.  A wrinkle here is that rounding errors
caused the two algorithms to give slightly different results.  

We also ran experiments to understand how best to choose parameters such as
the number of workers and the number of tasks.  The main mistake one can make
here is to make tasks  too small.  This leads to the channels becoming
bottlenecks, and the program being inefficient.  It can be best to run fewer
threads than the number of machine threads available, so as to avoid these
bottlenecks.  Each task should be large enough that workers spend nearly all
their time actually performing the tasks, as opposed to trying to send and
receive messages.

In Section~\ref{sec:bag-of-tasks-encapsulation}, we encapsulated the bag of
tasks and the collector inside objects.  This meant that we could concentrate
on that part of the implementation in isolation from other parts; and the
other parts of the program could use these objects without concern for how
they were implemented.  Further, this technique means that we could replace
those objects by other objects that produce the same external behaviour, but
perhaps using different concurrency techniques \framebox{forward ref here}.

Encapsulation like this will be an important theme in later chapters.  We will
arrange for nearly all of the concurrency to be encapsulated within such
objects.  This will make it much easier to justify correctness: local
correctness arguments that depend on a fairly small amount of code are easier
than global arguments that encompass a large program.  The correctness
property of such objects will be that the operation calls appear to take place
in a one-at-a-time order, without interfering with one another.

A variant on the bag-of-tasks pattern allows new tasks to be placed back into
the bag.  This can be the case where some tasks can be solved directly, but
other tasks lead to two or more recursive subtasks.  The latter case can be
tackled by returning subtasks to the bag.  We will need some extra machinery
to support returning tasks to the bag, which we will see in
Chapter~\ref{chap:alts}.

%% \emph{Adaptive quadrature} is an alternative approach that proceeds as
%% follows.  In order to calculate the integral $\int_a^b f(x) \mbox{d}x$,
%% compute the midpoint $mid = (a+b)/2$ and estimate three integrals, from $a$ to
%% $mid$, from $mid$ to~$b$, and from $a$ to~$b$, each using the trapezium rule
%% with a single interval.  If the sum of the former two estimates is within some
%% value $\epsilon$ of the third, then we take that third estimate as being the
%% result.  Otherwise, recursively estimate the integrals over the two
%% sub-intervals $a$ to~$mid$ and $mid$ to~$b$, and sum the results.

%% The latter case can be implemented by returning the two sub-intervals $(a,
%% mid)$ and $(mid, b)$ to the bag.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Exercises}

\input{Exercises/pi.tex}

\input{Exercises/matrix} % Matrix multiplication Q (sheet 1.5)

\framebox{?} Count \# common values (sheet 1.6) --- I think the best way to
solve this problem is to use a shared hash table; but that will appear later
in the book. 

\framebox{?} Maximum in array.




