\begin{question}
Implement a map, with the same interface as in Exercise~\ref{ex:sharded-map},
using a linked list with fine-grained locking.  Give you implementation the
following signature
\begin{scala}
class LinkedListMap[K,V] (implicit ord: K => Ordered[K]) extends Map[K,V]
\end{scala}
%
The linked list should be ordered by the keys (type~|K|).

Test your implementation using the linearizability testing framework. 
\end{question}

%%%%%

\begin{answerI}
My code is below.  The definitions of nodes and the |find| function are
straightforward adaptations of those for a sharded set.  Each of the main
operations requires a little thought.  The |key| field of |n = p.next| is
immutable, so it is not necessary to lock~|n| to access it.  However, it is
necessary to lock~|n| to obtain its |value| field.  Of course, we need to
remember to unlock relevant nodes in each branch. 

%%%%%

\begin{scala}
/** An implementation of a £Map£, using a linked list. */
class LinkedListMap[K,V] (implicit ord: K => Ordered[K]) extends Map[K,V]{
  /** Nodes from which the linked list is constructed.
    * Any access to the £next£ field must be done while holding the lock on this 
    * node.  */
  private class Node(val key: K, var value: V, var next: Node){
    /** A lock used to protect the £next£ field. */
    private val l = new Lock

    /** Lock this node. */
    def lock() = l.acquire()

    /** Unlock this node. */
    def unlock() = l.release()
  }

  /** A dummy header node. */
  private val header = new Node(null.asInstanceOf[K], null.asInstanceOf[V], null)

  /* Datatype invariant: the nodes from £header.next£ onwards are in strictly
   * increasing order of £key£ fields. */ 

  /** Find the first node £p£ that such that £p.next = null£ or £p.next.key $\ge$ k£.
    * £p£ will be locked on return. */  
  private def find(k: K): Node = {
    var p = header; p.lock()
    // Invariant: all nodes £n£ before £p£ have £n.next.key $<$ k£; £p£ is locked.
    while(p.next != null && p.next.key < k){
      val n = p.next; n.lock(); p.unlock(); p = n
    }
    p
  }

  /** Optionally get the value associated with £k£. */
  def get(k: K): Option[V] = {
    val p = find(k); val n = p.next
    if(n != null && n.key == k){ 
      n.lock(); p.unlock(); val res = n.value; n.unlock(); Some(res) 
    }
    else{ p.unlock(); None }
  }

  /** Add the association £k $\rightarrow$ v£.  Optionally return the value previously
    * associated with £k£. */
  def put(k: K, v: V): Option[V] = {
    val p = find(k); val n = p.next
    if(n != null && n.key == k){
      n.lock(); p.unlock(); val res = n.value; n.value = v; n.unlock(); Some(res)
    }
    else{ p.next = new Node(k, v, n); p.unlock(); None }
  }

  /** Remove any association for £k£.  Optionally return the value previously
    * associated with £k£. */
  def remove(k: K): Option[V] = {
    val p = find(k); val n = p.next
    if(n != null && n.key == k){
      n.lock(); p.next = n.next; p.unlock(); Some(n.value)
    }
    else{ p.unlock(); None }
  }
}
\end{scala}

Testing can be done as in Exercise~\ref{ex:sharded-map}.
\end{answerI}
